Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking

### 2023-07-12 manuscript/manuscript.tex
van hees spt chatter

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{lineno}\linenumbers
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{npj Digital Medicine}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Estimating Sleep Quality Metrics Using Free-Living Accelerometer Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep Tracking Device},
  pdfauthor={Esben Høegholm Lykke; Jan Christian Brønd},
  pdfkeywords={Sleep, Accelerometry, EEG, Machine learning, Sleep
quality},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Estimating Sleep Quality Metrics Using Free-Living Accelerometer
Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep
Tracking Device}
\author[1]{Esben Høegholm Lykke%
\corref{cor1}%
}
 \ead{eskovgaard@health.sdu.dk} 
\author[1]{Jan Christian Brønd%
%
}
 \ead{jbrond@health.sdu.dk}

@@

\begin{keyword}
    Sleep \sep Accelerometry \sep EEG \sep Machine learning \sep 
    Sleep quality
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, enhanced, interior hidden, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

In this study, we employed a biLSTM, a multiclass classifier, to predict
three distinct states: out-of-bed-awake, in-bed-awake, and
in-bed-asleep. The architecture of the biLSTM was set up with four
layers, each equipped with 128 hidden units. This configuration was
intentionally chosen to balance between model complexity and training
efficiency: it provided the depth necessary for learning intricate
patterns while remaining feasible for timely training. The bidirectional
design of the LSTM served to enhance data interpretation and mitigate
overfitting by doubling the hidden units at each time step. For input,
we used tensors shaped as sequences, with each sequence spanning 10
minutes and a step size of one.This approach follows in the footsteps of
previous studies that utilized LSTM models for sleep detection. These
studies showcased the promising potential of LSTMs in capturing complex
temporal patterns. Particularly, the works of Sano et al.
\citep{sano2019} and Chen et al. \citep{chen2021} demonstrated the
effectiveness of LSTM models in improving sleep detection using
accelerometer data, underscoring the value of this modeling approach.

@@

In the context of our sequential learning strategy, the initial models
were tasked with the binary classification of in-bed vs.~out-of-bed. For
this task, we assessed performance using the F1-score, accuracy,
sensitivity, specificity, and precision metrics. The second models in
our sequential learning strategy focused on the binary classification of
asleep vs.~awake. For these models, we considered the same metrics, in
addition to the negative predictive rate. The class imbalance in this
case led us to compute the F1 score as an unweighted macro-average.
Additionally, we evaluated the multiclass classifier, biLSTM, using
macro-averaged F1-score, sensitivity, and precision. To further
illustrate model performance, we provide confusion matrices for the full
dataset, encompassing both in-bed and out-of-bed data. These matrices
report relative counts, column percentages (the proportion of the true
class accurately predicted), and row percentages (the proportion of
predictions correctly classified). We considered both the
in-bed/out-of-bed and awake/asleep scoring tasks as binary
classification problems, designating in-bed and asleep as the positive
labels and out-of-bed and awake as the negative labels in accordance
with previous research\citep{hjorth2012, kushida2001}.

To assess the performance of our models in deriving sleep quality
parameters, we utilized Bland-Altman plots and Pearson correlations. The
Bland-Altman method was employed specifically to determine the level of
agreement between two measurement techniques. Considering our dataset
contained multiple observations per subject, we integrated a bootstrap
procedure to address this extra source of variability. We calculated the
mean difference (bias) and defined the limits of agreement as the mean
difference plus or minus 1.96 times the standard deviation of these
differences. To ensure our measurements were robust and accounted for
intra-subject variability, we estimated the 95\% confidence intervals
for both the bias and the limits of agreement using a bias-corrected and
accelerated bootstrap method, utilizing 10,000 bootstrap replicates. The
sleep quality parameteres included the following:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Sleep Period Time (SPT) - This refers to the total duration of the
  sleep period, which is defined as the time from the start to the end
  of the ZM recording.
\item
  Total Sleep Time (TST) - This is the time spent asleep within the SPT.
\item
  Sleep Efficiency (SE) - This is the ratio between TST and SPT,
  representing the proportion of the sleep period that was actually
  spent asleep.
\item
  Latency Until Persistent Sleep (LPS) - This metric represents the time
  it takes to transition from wakefulness to sustained sleep. It is
  calculated as the time from the beginning of the ZM recording until a
  period when 10 out of 12 minutes are scored as sleep.
\item
  Wake After Sleep Onset (WASO) - This refers to the time spent awake
  after initially falling asleep and before the final awakening. In our
  analysis, a period is counted as `awake' only if it consists of 3 or
  more contiguous 30-second epochs which is also how the ZM summarizes
  WASO.
\end{enumerate}

@@

\hypertarget{tbl-in_bed_performance}{}
\begin{longtable}{lrrrrr}
\caption{\label{tbl-in_bed_performance}In-Bed Performance Metrics }\tabularnewline

\toprule
 & F1 Score (\%) & Accuracy (\%) & Sensitivity (\%) & Precision (\%) & Specificity (\%) \\ 
\midrule
Decision Tree & $94.4$ & $95.3$ & $93.1$ & $95.6$ & $96.9$ \\ 
Logistic Regression & $95.0$ & $95.7$ & $95.0$ & $94.9$ & $96.3$ \\ 
Feed-Forward Neural Net & $95.0$ & $95.8$ & $95.1$ & $95.0$ & $96.3$ \\ 
XGBoost & $95.4$ & $96.1$ & $95.8$ & $94.9$ & $96.2$ \\ 
\bottomrule
\end{longtable}

Table~\ref{tbl-sleep_performance} details the performance of the second
sequential models on raw and median-filtered (5 and 10 minute) ZM
predictions. These are the predictions of the sequential models' ability
to predict asleep/awake on only the extracted in-bed time. The F1
scores, which are unweighted macro averages, for raw ZM predictions
range from 71.05\% to 76.18\%. The models perform comparably, but the
low specificity values (62.84\% to 70.93\%) suggest difficulty in
correctly classifying awake epochs. Applying 5-minute median filtering
improves the performance metrics. The XGBoost model tops the charts with
an F1 score of 79.22\% and NPV of 74.00\%. However, specificity still
remains low, with values between 54.68\% and 74.84\% across all models.
With 10-minute median filtering, the metrics improve further. The
XGBoost model still leads with an F1 score of 80.87\% and an NPV of
75.76\%. But, specificity remains low, ranging from 57.47\% to 76.35\%
across all models.

\hypertarget{tbl-sleep_performance}{}
\begin{longtable}{lrrrrr}
\caption{\label{tbl-sleep_performance}Sleep Performance Metrics }\tabularnewline

\toprule
 & F1 Score (\%) & Precision (\%) & NPV (\%) & Sensitivity (\%) & Specificity (\%) \\ 
\midrule
\multicolumn{6}{l}{Raw ZM Predictions} \\ 
\midrule
Decision Tree & $72.9$ & $93.2$ & $48.4$ & $86.3$ & $67.1$ \\ 
Logistic Regression & $71.0$ & $93.7$ & $43.9$ & $82.7$ & $70.9$ \\ 
Neural Network & $71.8$ & $93.8$ & $45.1$ & $83.6$ & $70.8$ \\ 
XGBoost & $76.2$ & $92.8$ & $58.0$ & $91.3$ & $62.8$ \\ 
\midrule
\multicolumn{6}{l}{5-Min Median} \\ 
\midrule
Decision Tree & $75.5$ & $94.2$ & $55.5$ & $93.4$ & $59.0$ \\ 
Logistic Regression & $68.3$ & $95.8$ & $36.0$ & $81.4$ & $74.8$ \\ 
Neural Network & $71.7$ & $95.8$ & $41.6$ & $85.6$ & $73.1$ \\ 
XGBoost & $79.2$ & $93.9$ & $74.0$ & $97.3$ & $54.7$ \\ 
\midrule
\multicolumn{6}{l}{10-Min Median} \\ 
\midrule
Decision Tree & $76.3$ & $94.7$ & $58.1$ & $94.9$ & $57.5$ \\ 
Logistic Regression & $68.0$ & $96.5$ & $34.3$ & $81.9$ & $76.4$ \\ 
Neural Network & $71.0$ & $96.1$ & $39.5$ & $86.5$ & $71.4$ \\ 
XGBoost & $80.9$ & $94.9$ & $75.8$ & $97.7$ & $57.6$ \\ 
\bottomrule
\end{longtable}

The complete set of confusion matrices generated from data both
containing the out-of-bed and in-bed time are presented in
Figure~\ref{fig-conf_mat}. The figure shows favorable epoch-to-epoch
performance across across all sequential models, however, it is evident
that the biLSTM is less successful in classifying the in-bed-awake class
which cannot be deduced from the confusion matrices from the sequential
models. \textbf{{[}fejl i biLSTM confmat. Alle mats er ens{]}}

@@

\caption{\label{fig-conf_mat}Confusion matrices for binary sleep
prediction. The middle of each tile is the normalized count (overall
percentage) and, beneath it, the count. The bottom number is the column
percentage (target). At the right side of each tile is the row
percentage (prediction). i) decision tree, ii) logistic regression, iii)
feed-forward neural net, iv) xgboost, and v) biLSTM.}

@@

Table~\ref{tbl-biLSTM_performance} presents the performance of the
three-class biLSTM multiclassifier on raw and median-filtered (5-minute
and 10-minute) ZM predictions. Raw ZM predictions achieve F1 Scores
ranging from 71.36\% to 76.04\%, indicating overall good performance.
Applying 5-minute median filtering improves the metrics further,
resulting in F1 Scores ranging from 75.99\% to 78.53\%, demonstrating
enhanced precision and sensitivity. \textbf{{[}Overvej flere metrics.
Evt. per class\ldots{} pt er det lidt svært at sammenligne på tværs af
lstm og andre modeller ud over F1.{]}} With 10-minute median filtering,
F1 Scores range from 73.45\% to 73.93\%, maintaining good performance
but with limited information on Specificity. Further analysis is
required to assess performance across all classes.

\hypertarget{tbl-biLSTM_performance}{}
\begin{longtable}{cccccc}
\caption{\label{tbl-biLSTM_performance}Performance of the three-class biLSTM multicclassifier. }\tabularnewline

\toprule
 & F1 Score & Precision & NPV & Sensitivity & Specificity \\ 
\midrule
Raw ZM Predictions & $71.4$ & $76.0$ & $95.5$ & $70.4$ & $95.0$ \\ 
5-Min Median & $76.0$ & $78.5$ & $96.4$ & $74.6$ & $96.2$ \\ 
10-Min Median & $73.4$ & $73.1$ & $95.6$ & $73.9$ & $95.7$ \\ 
\bottomrule
\end{longtable}

@@

Table~\ref{tbl-ba_cor} presents a comparative analysis of the included
models used to predict various sleep quality parameters (SPT, TST, SE,
LPS, WASO) using the 5-minute median filtered ZM predictions. To see the
full table including models developed from raw ZM predictions and
10-minute median filtered ZM predictions, see \textbf{{[}SUPP. MAT.{]}}.
In terms of bias, the decision tree model consistently underestimated
SPT, TST, and SE, and overestimated LPS and WASO in comparison to ZM.
The logistic regression model had similar trends, with more pronounced
underestimation in TST and overestimation in LPS. The eed-forward neural
network also exhibited similar bias as the decision tree and the
logistic regression models, but with a higher overestimation in WASO. On
the other hand, the XGboost model showed least bias among all,
especially in its 5-minute median predictions. Considering LOA, the
decision tree had higher variability across different sleep quality
parameters and filtering techniques, particularly for LPS and WASO,
which indicates lower agreement with ZM. Other models had comparable LOA
but with notable exceptions. For example, TST LOA for the logistic
regression model was particularly wide in the 5-minute median
predictions. Correlation-wise, the pearson coefficient, revealed that
the XGboost model consistently had the highest correlation with ZM
across all sleep qualityparameters and filtering methods Notably, the
XGboost's 5-minute median predictions showed the strongest correlation
(0.66) for TST among all models and filtering techniques.

\hypertarget{tbl-ba_cor}{}
\begin{longtable}{lrrrr}
\caption{\label{tbl-ba_cor}Summary of Bias, Limits of Agreement (LOA), and Pearson Correlation for
various Sleep Parameter Predictions (SPT, TST, SE, LPS, WASO) using
different Machine Learning Models (Decision Tree, Logistic Regression,
Feed-Forward Neural Net, XGBoost) with Raw ZM Predictions, 5-Min and
10-Min Median as predictors. Each value is provided with its 95\%
Confidence Interval (CI). }\tabularnewline

@@

Figure~\ref{fig-xgb_ba_cor} shows the agreement of the XGBoost model
trained on the 5-minute median filtered ZM predictions in estimating
sleep quality parameters as evaluated 5-minute median-smoothed ZM
derived sleep quality parameter. As can be seen on from the Bland-Altman
plots, a large portion of the nights are in almost perfect agreement
with the ZM and also a portion that can be considered outliers. This is
true for all sleep quality parameters.

\textbf{{[}Overvej at smide samtlige 75 plots i supp.mat.{]}}

@@

\caption{\label{fig-xgb_ba_cor}Comparison of sleep quality parameters
derived from the XGboost model trained on the 5-minute smoothed ZM
predictions. The left column displays Bland-Altman plots. Dashed lines
represent the bias (the average difference between the two measurements)
and limits of agreement, with the 95\% confidence intervals represented
as the grayed areas. The right column displays scatter plots of
XGboost-derived vs ZM-derived sleep quality parameters. The dashed line
represents the identity line, while the full-drawn line represents the
best linear fit. Pearson's correlations are annotated in the upper left
corner.}

@@

The present study embarked on an in-depth exploration of the utility and
comparative performance of different machine learning models in the
analysis of pediatric sleep data. The results offer promising insights
into the role of machine learning in enhancing the precision of sleep
quality metrics.

A significant finding was the impact of median filters on the sleep
metrics. Implementing 5-minute and 10-minute median filters led to
notable changes in key sleep parameters. In particular, the Total Sleep
Time (TST), Sleep Efficiency (SE), and Latency to Persistent Sleep (LPS)
increased, suggesting that the filters may categorize some instances of
wakefulness as sleep and smooth out brief awakenings. This underscores
the potential of median filters in refining sleep data and mitigating
the potential misclassification of sleep states.

Interestingly, the most marked change was observed in Wake After Sleep
Onset (WASO), with a substantial reduction in the filtered data. This
reduction, coupled with the decrease in the number of awakenings,
implies a significant role of median filters in sleep-wake cycle
characterization, potentially enhancing the reliability of sleep
assessment.

The machine learning models demonstrated high overall performance on an
epoch-to-epoch basis, with the XGBoost model marginally outperforming
the others. However, the models had more varied performance in
classifying sleep conditions, as evidenced by the sleep performance
metrics. The relatively lower Specificity values across all models
suggest a common challenge in correctly classifying negative instances.

In the precision-recall and ROC curve analyses, the Decision Tree model
showed superior precision-recall AUC, indicating strong predictive
accuracy. In contrast, the XGBoost model consistently demonstrated a
high ROC AUC, indicating a robust ability to differentiate between
classes. However, it's worth noting that the Neural Network model
generally showed weaker performance in these areas.

>>>>

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{lineno}\linenumbers
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{npj Digital Medicine}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking},
  pdfauthor={Esben Høegholm Lykke; Jan Christian Brønd},
  pdfkeywords={Sleep, Accelerometry, EEG, Machine learning, Sleep
quality},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Improving Sleep Quality Estimation: A Comparative Study of
Machine Learning and Deep Learning Techniques Utilizing Free-Living
Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking}
\author[1]{Esben Høegholm Lykke%
\corref{cor1}%
}
 \ead{eskovgaard@health.sdu.dk} 
\author[1]{Jan Christian Brønd%
%
}
 \ead{jbrond@health.sdu.dk}

@@

\begin{keyword}
    Sleep \sep Accelerometry \sep EEG \sep Machine learning \sep 
    Sleep quality
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden, enhanced, sharp corners, interior hidden]}{\end{tcolorbox}}\fi

@@

We also employed a biLSTM, a multiclass classifier, to predict three
distinct states: out-of-bed-awake, in-bed-awake, and in-bed-asleep. The
architecture of the biLSTM was set up with four layers, each equipped
with 128 hidden units. This configuration was intentionally chosen to
balance between model complexity and training efficiency: it provided
the depth necessary for learning intricate patterns while remaining
feasible for timely training. The bidirectional design of the LSTM
served to enhance data interpretation and mitigate overfitting by
doubling the hidden units at each time step. For input, we used tensors
shaped as sequences, with each sequence spanning 10 minutes and a step
size of one.This approach follows in the footsteps of previous studies
that utilized LSTM models for sleep detection. These studies showcased
the promising potential of LSTMs in capturing complex temporal patterns.
Particularly, the works of Sano et al. \citep{sano2019} and Chen et al.
\citep{chen2021} demonstrated the effectiveness of LSTM models in
improving sleep detection using accelerometer data, underscoring the
value of this modeling approach.

@@

In the context of our sequential learning strategy, the initial models
were tasked with the binary classification of in-bed vs.~out-of-bed. For
this task, we assessed performance using the F1-score, accuracy,
sensitivity, specificity, and precision metrics. The second models in
our sequential learning strategy focused on the binary classification of
asleep vs.~awake. For these models, we considered the same metrics, in
addition to the negative predictive rate. The class imbalance in this
case led us to compute the F1 score as an unweighted macro-average.
Additionally, we evaluated the multiclass classifier, biLSTM, using the
same metrics. To do this, we considered the multiclass output as to
binary classifications, where the first was out-of-bed vs the rest and
the second binary classification as in-bed-awake vs in-bed-asleep. To
further illustrate model performance, we provide confusion matrices for
the full dataset, encompassing both in-bed and out-of-bed data. These
matrices report relative counts, column percentages (the proportion of
the true class accurately predicted), and row percentages (the
proportion of predictions correctly classified). We considered both the
in-bed/out-of-bed and awake/asleep scoring tasks as binary
classification problems, designating in-bed and asleep as the positive
labels and out-of-bed and awake as the negative labels in accordance
with previous research\citep{hjorth2012, kushida2001}.

To assess the performance of our models in deriving sleep quality
parameters, we utilized Bland-Altman plots and Pearson correlations. The
Bland-Altman method was employed specifically to determine the level of
agreement between two measurement techniques. Considering our dataset
contained multiple observations per subject, we integrated a bootstrap
procedure to address this extra source of variability. We calculated the
mean difference (bias) and defined the LOA as the mean difference plus
or minus 1.96 times the standard deviation of these differences. To
ensure our measurements were robust and accounted for intra-subject
variability, we estimated the 95\% confidence intervals for both the
bias and the LOA using a bias-corrected and accelerated bootstrap
method, utilizing 10,000 bootstrap replicates. The sleep quality
parameteres included are defined as follews in accordance with the ZM
definitions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Sleep Period Time (SPT) - This refers to the total duration of the
  sleep period, which is defined as the time from the start to the end
  of the ZM recording.
\item
  Total Sleep Time (TST) - This is the time spent asleep within the SPT.
\item
  Sleep Efficiency (SE) - This is the ratio between TST and SPT,
  representing the proportion of the sleep period that was actually
  spent asleep.
\item
  Latency Until Persistent Sleep (LPS) - This metric represents the time
  it takes to transition from wakefulness to sustained sleep. It is
  calculated as the time from the beginning of the ZM recording until
  the first period when 10 out of 12 minutes are scored as sleep.
\item
  Wake After Sleep Onset (WASO) - This refers to the time spent awake
  after initially falling asleep and before the final awakening. In our
  analysis, a period is counted as `awake' only if it consists of 3 or
  more contiguous 30-second epochs which is also how the ZM summarizes
  WASO.
\end{enumerate}

@@

\hypertarget{tbl-in_bed_performance}{}
\begin{longtable}{lrrrrr}
\caption{\label{tbl-in_bed_performance}In-bed performance metrics }\tabularnewline

\toprule
 & F1 Score (\%) & Accuracy (\%) & Sensitivity (\%) & Precision (\%) & Specificity (\%) \\ 
\midrule
Decision Tree & $94.4$ & $95.3$ & $93.1$ & $95.6$ & $96.9$ \\ 
Logistic Regression & $95.0$ & $95.7$ & $95.0$ & $94.9$ & $96.3$ \\ 
Feed-Forward Neural Net & $95.0$ & $95.8$ & $95.1$ & $95.0$ & $96.3$ \\ 
XGBoost & $95.4$ & $96.1$ & $95.8$ & $94.9$ & $96.2$ \\ 
biLSTM & $95.2$ & $95.3$ & $95.3$ & $95.1$ & $95.3$ \\ 
\bottomrule
\end{longtable}

Table~\ref{tbl-sleep_performance} details the performance of all
sequential model types on raw and median-filtered (5 and 10 minute) ZM
predictions for sleep/wake classification. For raw ZM predictions, the
F1 scores, which are unweighted macro averages, range from 65.6\%
(biLSTM) to 76.2\% (XGBoost). All models perform comparably, but the low
specificity values (62.5\% to 70.9\%) suggest difficulty in correctly
classifying awake epochs. Applying a 5-minute median filter improves the
performance metrics. The XGBoost model tops the charts with an F1 score
of 79.2\% and NPV of 74.0\%. However, specificity still remains low,
with values between 54.7\% (XGBoost) and 74.8\% (Logistic Regression)
across all models. With a 10-minute median filter, the metrics improve
further. The XGBoost model still leads with an F1 score of 80.9\% and an
NPV of 75.9\%. But, specificity remains low, ranging from 57.5\%
(Decision Tree) to 76.4\% (Logistic Regression) across all models.

\hypertarget{tbl-sleep_performance}{}
\begin{longtable}{lrrrrr}
\caption{\label{tbl-sleep_performance}Performance of the sleep/wake classification of the sequential models. }\tabularnewline

\toprule
 & F1 Score (\%) & Precision (\%) & NPV (\%) & Sensitivity (\%) & Specificity (\%) \\ 
\midrule
\multicolumn{6}{l}{Raw ZM Predictions} \\ 
\midrule
Decision Tree & $72.9$ & $93.2$ & $48.4$ & $86.3$ & $67.1$ \\ 
Logistic Regression & $71.0$ & $93.7$ & $43.9$ & $82.7$ & $70.9$ \\ 
Neural Network & $71.8$ & $93.8$ & $45.1$ & $83.6$ & $70.8$ \\ 
XGBoost & $76.2$ & $92.8$ & $58.0$ & $91.3$ & $62.8$ \\ 
biLSTM & $65.6$ & $80.6$ & $80.6$ & $62.5$ & $62.5$ \\ 
\midrule
\multicolumn{6}{l}{5-Min Median} \\ 
\midrule
Decision Tree & $75.5$ & $94.2$ & $55.5$ & $93.4$ & $59.0$ \\ 
Logistic Regression & $68.3$ & $95.8$ & $36.0$ & $81.4$ & $74.8$ \\ 
Neural Network & $71.7$ & $95.8$ & $41.6$ & $85.6$ & $73.1$ \\ 
XGBoost & $79.2$ & $93.9$ & $74.0$ & $97.3$ & $54.7$ \\ 
biLSTM & $70.3$ & $84.6$ & $84.6$ & $66.2$ & $66.2$ \\ 
\midrule
\multicolumn{6}{l}{10-Min Median} \\ 
\midrule
Decision Tree & $76.3$ & $94.7$ & $58.1$ & $94.9$ & $57.5$ \\ 
Logistic Regression & $68.0$ & $96.5$ & $34.3$ & $81.9$ & $76.4$ \\ 
Neural Network & $71.0$ & $96.1$ & $39.5$ & $86.5$ & $71.4$ \\ 
XGBoost & $80.9$ & $94.9$ & $75.8$ & $97.7$ & $57.6$ \\ 
biLSTM & $70.9$ & $75.1$ & $75.1$ & $68.5$ & $68.5$ \\ 
\bottomrule
\end{longtable}

A complete set of confusion matrices generated from data both containing
the out-of-bed and in-bed time are presented in
Figure~\ref{fig-conf_mat}. The figure shows favorable epoch-to-epoch
performance across across all sequential models, however, it is evident
that the biLSTM is less successful in classifying the in-bed-awake class
which cannot be deduced from the confusion matrices from the sequential
models.

@@

\caption{\label{fig-conf_mat}Confusion matrices for binary sleep
prediction. The middle of each tile is the normalized count (overall
percentage) and, beneath it, the count. The bottom number is the column
percentage (target). At the right side of each tile is the row
percentage (prediction). i) decision tree, ii) logistic regression, iii)
feed-forward neural net, iv) XGBoost, and v) biLSTM.}

@@

@@

Table~\ref{tbl-ba_cor} presents a comparative analysis of the included
models used to predict various sleep quality parameters (SPT, TST, SE,
LPS, WASO) using the 5-minute median filtered ZM predictions. To see the
full table including models developed from raw ZM predictions and
10-minute median filtered ZM predictions, see \textbf{{[}SUPP. MAT.{]}}.
In terms of bias, the decision tree model consistently underestimated
SPT, TST, and SE, and overestimated LPS and WASO in comparison to ZM.
The logistic regression model had similar trends, with more pronounced
underestimation in TST and overestimation in LPS. The eed-forward neural
network also exhibited similar bias as the decision tree and the
logistic regression models, but with a higher overestimation in WASO. On
the other hand, the XGBoost model showed least bias among all,
especially in its 5-minute median predictions. Considering LOA, the
decision tree had higher variability across different sleep quality
parameters and filtering techniques, particularly for LPS and WASO,
which indicates lower agreement with ZM. Other models had comparable LOA
but with notable exceptions. For example, TST LOA for the logistic
regression model was particularly wide in the 5-minute median
predictions. Correlation-wise, the pearson coefficient, revealed that
the XGBoost model consistently had the highest correlation with ZM
across all sleep qualityparameters and filtering methods Notably, the
XGBoost's 5-minute median predictions showed the strongest correlation
(0.66) for TST among all models and filtering techniques.

\hypertarget{tbl-ba_cor}{}
\begin{longtable}{lrrrr}
\caption{\label{tbl-ba_cor}Summary of Bias, Limits of Agreement, and Pearson Correlation for
various Sleep Parameter Predictions (SPT, TST, SE, LPS, WASO) using
different Machine Learning Models (Decision Tree, Logistic Regression,
Feed-Forward Neural Net, XGBoost) with Raw ZM Predictions, 5-Min and
10-Min Median as predictors. Each value is provided with its 95\%
Confidence Interval (CI). }\tabularnewline

@@

Figure~\ref{fig-xgb_ba_cor} shows the agreement between the XGBoost
model, trained on 5-minute median filtered ZM predictions, and the
5-minute median-smoothed ZM-derived sleep quality parameters. The
Bland-Altman plot for the SPT reveals a significant level of agreement
with the ZM, as evidenced by a positive correlation. However, the
presence of extreme outliers widens the limits of agreement (LOA). The
scatterplot for SPT also demonstrates a positive trend, indicating a
moderate linear correlation between the XGBoost model and the ZM-derived
sleep quality parameters.In terms of TST, perfect agreement is not
observed for a substantial number of nights, but there is a positive
correlation between the XGBoost model and ZM-derived sleep quality
parameters. The bias and LOA for TST are comparable to those observed
for SPT, indicating a consistent level of agreement between the two
methods. The scatterplot for TST also shows a slightly higher
correlation, primarily driven by the absence of extreme
outliers.Furthermore, the remaining three sleep quality parameters, SE,
LPS, and WASO, exhibit heteroscedasticity in contrast to SPT and TST.
This outcome is expected as achieving 100\% sleep efficiency is
relatively rare, resulting in less disagreement between the methods as
values approach the upper limit. However, as sleep efficiency decreases,
the potential for discrepancies and differing interpretations between
the methods increases, leading to greater heteroscedasticity. A moderate
linear correlation is observed between the XGBoost model and ZM-derived
sleep quality parameters for SE, indicating a positive relationship.
However, a poor correlation is observed for LPS and WASO, suggesting
less agreement between the two methods for these parameters. Similar
plots for all models are available in the supplementary materials.

@@

\caption{\label{fig-xgb_ba_cor}Comparison of sleep quality parameters
derived from the XGBoost model trained on the 5-minute smoothed ZM
predictions. The left column displays Bland-Altman plots. Dashed lines
represent the bias (the average difference between the two measurements)
and LOA, with the 95\% confidence intervals represented as the grayed
areas. The right column displays scatter plots of XGBoost-derived vs
ZM-derived sleep quality parameters. The dashed line represents the
identity line, while the full-drawn line represents the best linear fit.
Pearson's correlations are annotated in the upper left corner.}

@@

In an effort to mature the methods for estimating sleep from thigh-worn
accelerometers, we evaluated various models for predicting in-bed and
sleep time and their derived sleep quality parameters. Furthermore, we
trained and evaluated the models using raw and median-filtered gold
standard predictions from the ZM. In general, all sequential models
performed well at predicting in-bed time. More challenging was it to
distinguish wake from sleep on the extracted in-bed time, however, the
performance of the sequential models were enhanced by the application of
median filters. Moreover, even though the multiclass biLSTM showed good
performance across F1 score, precision and NPV, the derived sleep
quality parameters were not on par with the XGBoost model which
demonstrated the highest performance metrics across all evaluations,
including epoch-to-epoch prediction and sleep quality parameter
derivatives. Despite this, all sequential models showed low specificity
values, indicating difficulty in correctly classifying awake epochs
during time in bed. The application of 5-minute and 10-minute median
filters improved the performance metrics of all models, with the XGBoost
model consistently leading. The median filters increased total sleep
time and sleep efficiency, while reducing wake after sleep onset and the
number of awakenings. The XGBoost model also showed the least bias and
highest correlation with ZM across sleep quality parameters.
